---
title: "Hydrochemical Data Analysis"
subtitle: "PCA, Factor Analysis, and Visualization of Spa Water Chemistry"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    embed-resources: true
execute:
  warning: false
  message: false
---

## Setup

Install packages (run once, then comment out).

```{r}
#| label: install-packages
#| eval: false

install.packages(c("reshape2", "ggplot2", "ggcorrplot", "GGally", "psych", "ggnewscale"))
install.packages("devtools")
devtools::install_github("vqv/ggbiplot")
```

Load libraries.

```{r}
#| label: load-libraries

library(reshape2)
library(ggplot2)
library(ggcorrplot)
library(GGally)
library(ggbiplot)
library(psych)
library(ggnewscale)
```

## Data Preparation

### Reading the data

```{r}
#| label: read-data

data_spa_waters <- read.csv("C:\\Users\\ixion\\NZ\\PhD\\STAT394_Group_6\\Data\\Hydrochemical Data.csv") # Modify this path
head(data_spa_waters)
```

### Cleaning data

Remove the first row and the first 2 columns (number, place name).

```{r}
#| label: clean-data

data_spa_waters_classes <- data_spa_waters[-1, -c(1, 2)]
ncolumns_data_spa_waters_classes <- ncol(data_spa_waters_classes)
```

Consider data as numeric, round to 3 decimals except the class columns.

```{r}
#| label: convert-numeric

data_spa_waters_classes[-c(ncolumns_data_spa_waters_classes - 1, ncolumns_data_spa_waters_classes)] <-
  lapply(
    data_spa_waters_classes[-c(ncolumns_data_spa_waters_classes - 1, ncolumns_data_spa_waters_classes)],
    function(x) round(as.numeric(x), 3)
  )

# Set Structure and Region as factors
data_spa_waters_classes$Geological.structure <- as.factor(data_spa_waters_classes$Geological.structure)
data_spa_waters_classes$Region <- as.factor(data_spa_waters_classes$Region)
```

Show the data.

```{r}
#| label: show-classes-data

head(data_spa_waters_classes)
summary(data_spa_waters_classes)
```

Remove the last 2 columns to get the raw numerical data.

```{r}
#| label: raw-data

data_spa_waters_raw <- data_spa_waters_classes[, -c(ncolumns_data_spa_waters_classes - 1, ncolumns_data_spa_waters_classes)]

head(data_spa_waters_raw)
summary(data_spa_waters_raw)
```

## Correlation Analysis

```{r}
#| label: correlation-matrix

cor(data_spa_waters_raw)
```

### Pairs plot

```{r}
#| label: pairs-plot
#| fig-width: 9
#| fig-height: 9

ggpairs(data_spa_waters_raw)
```

### Correlogram

```{r}
#| label: correlogram

ggcorrplot(cor(data_spa_waters_raw),
            method = "circle",
            type = "lower",
            lab = TRUE)
```

### Correlogram, hierarchically ordered

```{r}
#| label: correlogram-ordered

ggcorrplot(cor(data_spa_waters_raw),
            method = "circle",
            hc.order = TRUE,
            type = "lower",
            lab = TRUE)
```

> **TODO:** analysing correlations between compounds

## Multivariate Outlier Assessment (Mahalanobis Distance)

Mahalanobis distance for assessing measurements multivariately.

```{r}
#| label: mahalanobis-setup

(mu.hat <- colMeans(data_spa_waters_raw))
(Sigma.hat <- cov(data_spa_waters_raw))

dM <- mahalanobis(data_spa_waters_raw, center = mu.hat, cov = Sigma.hat)

upper.quantiles <- qchisq(c(.9, .95, .99), df = 9)
density.at.quantiles <- dchisq(x = upper.quantiles, df = 9)
cut.points <- data.frame(upper.quantiles, density.at.quantiles)
```

```{r}
#| label: mahalanobis-plot

ggplot(data.frame(dM), aes(x = dM)) +
  geom_histogram(aes(y = after_stat(density)), bins = nclass.FD(dM),
                 fill = "white", col = "black") +
  geom_rug() +
  stat_function(fun = dchisq, args = list(df = 9),
                col = "red", size = 2, alpha = .7, xlim = c(0, 25)) +
  geom_segment(data = cut.points,
               aes(x = upper.quantiles, xend = upper.quantiles,
                   y = rep(0, 3), yend = density.at.quantiles),
               col = "blue", size = 2) +
  xlab("Mahalanobis distances and cut points") +
  ylab("Histogram and density")
```

### Assessing measurements

```{r}
#| label: surprise-classification

data_spa_waters_raw$dM <- dM
data_spa_waters_raw$surprise <- cut(
  data_spa_waters_raw$dM,
  breaks = c(0, upper.quantiles, Inf),
  labels = c("Typical", "Somewhat", "Surprising", "Very")
)
table(data_spa_waters_raw$surprise)
```

### Pairs plot colored by surprise level

```{r}
#| label: pairs-plot-surprise
#| fig-width: 9
#| fig-height: 9

ggpairs(data_spa_waters_raw, columns = 1:9,
        ggplot2::aes(col = surprise, alpha = .5),
        upper = list(continuous = "density", combo = "box_no_facet")) +
  ggplot2::scale_color_manual(values = c("lightgray", "green", "blue", "red")) +
  ggplot2::theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## Analyzing the Data with Classes

### Melting the data

```{r}
#| label: melt-data

data_spa_waters_classes.melt <- melt(
  data = data_spa_waters_classes[-ncolumns_data_spa_waters_classes],
  id.vars = "Geological.structure",
  variable.name = "Chemical_Compound"
)
```

### Box plots with notches

```{r}
#| label: boxplot-notched
#| fig-width: 9

ggplot(data = data_spa_waters_classes.melt, aes(x = Chemical_Compound, y = value)) +
  geom_boxplot(aes(col = Chemical_Compound), notch = TRUE) +
  facet_grid(~ Geological.structure) +
  ylab("Measurements") +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        legend.position = "none")
```

> Notice, the data doesn't have enough samples for the notches to appear nicely.

### Box plots, log scale geology

```{r}
#| label: boxplot-log
#| fig-width: 9

ggplot(data = data_spa_waters_classes.melt, aes(x = Chemical_Compound, y = value)) +
  geom_boxplot(aes(col = Chemical_Compound), notch = FALSE) +
  scale_y_log10() +
  facet_grid(. ~ Geological.structure) +
  ylab("Measurements") +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        legend.position = "none")
```

### Box plots, log scale region
```{r}
#| label: melt-data

data_spa_waters_classes.melt <- melt(
  data = data_spa_waters_classes[-(ncolumns_data_spa_waters_classes-1)],
  id.vars = "Region",
  variable.name = "Chemical_Compound"
)
```

```{r}
#| label: boxplot-log
#| fig-width: 9

ggplot(data = data_spa_waters_classes.melt, aes(x = Chemical_Compound, y = value)) +
  geom_boxplot(aes(col = Chemical_Compound), notch = FALSE) +
  scale_y_log10() +
  facet_grid(. ~ Region) +
  ylab("Measurements") +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        legend.position = "none")
```


> **TODO:** Analyse box plots

## Principal Component Analysis (PCA)

Reset `data_spa_waters_raw` back to the original (without the Mahalanobis columns added earlier).

```{r}
#| label: pca-reset-raw

data_spa_waters_raw <- data_spa_waters_classes[, -c(ncolumns_data_spa_waters_classes - 1, ncolumns_data_spa_waters_classes)]
```

### Calculate PCA

```{r}
#| label: pca-calculate

PCA.data_spa_waters_raw <- prcomp(data_spa_waters_raw, center = TRUE, scale = TRUE)
summary(PCA.data_spa_waters_raw)
```

### Variance explained (scree plot)

```{r}
#| label: pca-scree

plot(PCA.data_spa_waters_raw, type = "l")
```

> **TODO:** Say some words about it

### PCA components

```{r}
#| label: pca-components

options(digits = 4)
PCA.data_spa_waters_raw$center   # Means of the data
PCA.data_spa_waters_raw$sdev     # Standard deviations of the data
PCA.data_spa_waters_raw$rotation # PC contributions of each variable (eigenvectors)
```

> **TODO:** Some words about it

### Biplots

Biplot with unit circle (PC1 vs PC2).

```{r}
#| label: biplot-circle

ggbiplot(PCA.data_spa_waters_raw, circle = TRUE)
```

Biplot grouped by Geological structure.

```{r}
#| label: biplot-geological-structure

ggbiplot(PCA.data_spa_waters_raw, obs.scale = 1, var.scale = 1,
         groups = interaction(data_spa_waters_classes$Geological.structure), ellipse = TRUE)
```

Biplot grouped by Region.

```{r}
#| label: biplot-region

ggbiplot(PCA.data_spa_waters_raw, obs.scale = 1, var.scale = 1,
         groups = interaction(data_spa_waters_classes$Region), ellipse = TRUE)
```

```{r}
#| label: pca-scores-summary

summary(PCA.data_spa_waters_raw$x)
```

Biplot for PC1 vs PC3.

```{r}
#| label: biplot-pc1-pc3

ggbiplot(PCA.data_spa_waters_raw, obs.scale = 1, var.scale = 1, circle = TRUE, choices = c(1, 3))

ggbiplot(PCA.data_spa_waters_raw, obs.scale = 1, var.scale = 1, choices = c(1, 3),
         groups = interaction(data_spa_waters_classes$Geological.structure), ellipse = TRUE)

ggbiplot(PCA.data_spa_waters_raw, obs.scale = 1, var.scale = 1, choices = c(1, 3),
         groups = interaction(data_spa_waters_classes$Region), ellipse = TRUE)
```

> **TODO:** Analyse the pair plots

```{r}
#| label: pca-importance

print(summary(PCA.data_spa_waters_raw)$importance[, 1:4])
```

## Factor Analysis

### Factor analysis with `factanal`

Choosing number of factors via p-value.

```{r}
#| label: factanal-pvalues

n <- 4
pvalues <- rep(0, n)

for (f in 1:n)
  pvalues[f] <- factanal(data_spa_waters_raw, factors = f)$PVAL

ggplot(data.frame(pvalues), aes(x = 1:n, y = pvalues)) +
  geom_point(size = 3) +
  xlab("Number of factors")
```

```{r}
#| label: factanal-4factors

factanal(data_spa_waters_raw, factors = 4)
```

```{r}
#| label: factanal-diagram

fa.diagram(factanal(data_spa_waters_raw, factors = 4)$loadings, digits = 3)
```

```{r}
#| label: factanal-scores

fa_result <- factanal(data_spa_waters_raw, factors = 4, rotation = "varimax", scores = "regression")
fa_result$scores
```

```{r}
#| label: factanal-scores-plot
#| eval: false

# Note: only run if fa_result$scores is not NULL
ggplot(fa_result$scores, aes(x = Factor1, y = Factor2)) +
  geom_point()
```

### Factor Analysis based on Principal Components (Varimax rotation)

```{r}
#| label: principal-varimax

fa_result <- principal(data_spa_waters_raw, nfactors = 4, rotate = "varimax")

# Check variance accounted for by each rotated factor
fa_result$Vaccounted

fa_result$loadings

fa.diagram(principal(data_spa_waters_raw, nfactors = 4, rotate = "varimax")$loadings, digits = 3)
```

## Score Plots

## Clustering on PCA Scores

```{r}
#| label: clustering-kmeans

# 1. Build the scores data frame
scores_df <- as.data.frame(PCA.data_spa_waters_raw$x)
scores_df$Group <- data_spa_waters_classes$Geological.structure #Region
scores_df$Label <- 1:nrow(scores_df)

# 2. Variance explained for axis labels
var_explained <- (PCA.data_spa_waters_raw$sdev^2 / sum(PCA.data_spa_waters_raw$sdev^2)) * 100

# 3. Run clustering on the PC scores (choose number of clusters, e.g. 3)
set.seed(123)
k <- 2  # adjust based on your data
km <- kmeans(scores_df[, c("PC1", "PC2")], centers = k)
scores_df$Cluster <- factor(km$cluster)

# 4. Dynamic color/shape mapping for Group
groups <- unique(scores_df$Group)
my_colors <- setNames(c("magenta", "blue", "black", "red", "darkgreen", "orange")[1:length(groups)], groups)
my_shapes <- setNames(c(8, 16, 0, 2, 17, 15)[1:length(groups)], groups)

ggplot(scores_df, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(group = Cluster), type = "norm", level = 0.95,
               color = "grey30", linetype = 2, linewidth = 0.8) +
  geom_point(aes(color = Group, shape = Group), size = 3, stroke = 1.2) +
  geom_text(aes(label = Label, color = Group), vjust = -0.8, hjust = 0.5,
            size = 3.5, fontface = "bold", show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = my_colors) +
  scale_shape_manual(values = my_shapes) +
  labs(x = paste0("Factor 1 (PC1) : ", round(var_explained[1], 2), "%"),
       y = paste0("Factor 2 (PC2) : ", round(var_explained[2], 2), "%")) +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = c(0.2, 0.85),
        legend.background = element_rect(color = "black", linewidth = 0.3))
```
# Doing it with Factors and see
