---
title: "mvnreport"
author: "Jack Hardy"
date: "2026-09-12"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
hydrodata<-read.csv("/Users/jackhardy/Documents/GitHub/STAT394_Group_6/Data/Hydrochemical Data.csv")

summary(hydrodata)

hydroclean<-hydrodata[-1, 3:14]
as.numeric(hydroclean$T)->hydroclean$T
as.numeric(hydroclean$EC)->hydroclean$EC
as.numeric(hydroclean$TS)->hydroclean$TS
as.numeric(hydroclean$Ca2.)->hydroclean$Ca2.

head(hydroclean)
summary(hydroclean)
```

The multivariate normal distribution is a well-established and mathematically convenient model that underlies many statistical procedures. So it is a reasonable and important step to understand our data, a step which might decide our subsequent analysis. Tanaskovic et al. tested whether the individual variables followed a normal distribution before their analysis, and ultimately transformed the variables which failed the Shapiro-Wilk tests for normality ($H_0$: the data comes from a normally distributed population) using a Box-Cox transformation. It is simple to attempt this ourselves; we can even look at histograms and QQ-plots to visually see the same. 

```{r}
#Histograms
par(mfrow = c(2,2))
hist(hydroclean$T, breaks = 25)
hist(hydroclean$pH, breaks = 25)
hist(hydroclean$EC, breaks = 25)
hist(hydroclean$TS, breaks =25)
par(mfrow = c(2,2))
hist(hydroclean$Ca2., breaks =25)
hist(hydroclean$Mg2., breaks =25)
hist(hydroclean$Na., breaks =25)
hist(hydroclean$K., breaks =25)
par(mfrow = c(2,2))
hist(hydroclean$Cl., breaks =25)
hist(hydroclean$SO2.4, breaks =25)
hist(hydroclean$HCO.3, breaks =25)
hist(hydroclean$SiO2, breaks =25)

#QQ-plots
par(mfrow = c(2,2))
qqnorm(hydroclean$T, main = "T")
qqline(hydroclean$T)
qqnorm(hydroclean$pH, main = "pH")
qqline(hydroclean$pH)
qqnorm(hydroclean$EC, main = "EC")
qqline(hydroclean$EC)
qqnorm(hydroclean$TS, main = "TS")
qqline(hydroclean$TS)
par(mfrow = c(2,2))
qqnorm(hydroclean$Ca2., main = "Ca2.")
qqline(hydroclean$Ca2.)
qqnorm(hydroclean$Mg2., main = "Mg2.")
qqline(hydroclean$Mg2.)
qqnorm(hydroclean$Na., main = "Na.")
qqline(hydroclean$Na.)
qqnorm(hydroclean$K., main = "K.")
qqline(hydroclean$K.)
par(mfrow = c(2,2))
qqnorm(hydroclean$Cl., main = "Cl.")
qqline(hydroclean$Cl.)
qqnorm(hydroclean$SO2.4, main = "SO2.4")
qqline(hydroclean$SO2.4)
qqnorm(hydroclean$HCO.3, main = "HCO.3")
qqline(hydroclean$HCO.3)
qqnorm(hydroclean$SiO2, main = "SiO2")
qqline(hydroclean$SiO2)

#The output of the code above can be reduced by par(mfrow = c(3,4))

#Shapiro-Wilk Tests
shapiro.test(hydroclean$T)
shapiro.test(hydroclean$pH)
shapiro.test(hydroclean$EC)
shapiro.test(hydroclean$TS)
shapiro.test(hydroclean$Ca2.)
shapiro.test(hydroclean$Mg2.)
shapiro.test(hydroclean$Na.)
shapiro.test(hydroclean$K.)
shapiro.test(hydroclean$Cl.)
shapiro.test(hydroclean$SO2.4)
shapiro.test(hydroclean$HCO.3)
shapiro.test(hydroclean$SiO2)
```

The histograms and QQ-plots show the individual variables deviate from normality to varying degrees. Understandably, for what the variables are, we see the measures are bounded below by zero and generally right skewed, where most observations are relatively low with a few very high values. Obviously these characteristics are different from the unbounded, symmetric normal distribution. 

Just like Tanaskovic, our Shapiro-Wilk tests conclude non-normality for our variables. In fact, we find at the 0.05 level (the significance level chosen in the Tanaskovic et al. (2012)), we reject the null hypothesis for every variable. For example, the chlorine ion measure (Cl.) gets a p-value of $1.566\times10^{-10} \ll 0.05$  so we reject the null hypothesis. For the Calcium ion variable (Ca2.), the p-value is 0.02344. Therefore we reject the null hypothesis at the 0.05  significance level but we would fail to reject the null at the 0.01 significance level had we chosen it.

And to the big question: does our data follow a multivariate normal distribution? We can run a couple of tests for this very question (significance level 0.05). The first we ran is Mardia's test which formally tests whether multiple variables considered together follows a multivariate normal distribution. There are a few things we need to have for Mardia's test to work its best. First we want independent observations, we don't want extreme outliers, and also want a reasonable sample size. In this case, we should only worry about the size of our sample, 30 observations is definitely smaller. Mardia's test has the null hypothesis the data follows a multivariate normal distribution and looks at two features of the data: skewness and kurtosis. Essentially whether the overall distribution is asymetric, and whether the overall distribution has heavy or light tails.

```{r}
library(psych)
mardia(hydroclean)
```

The p-value for skewness is $4.3\times10^{-20}$ and the p-value for kurtosis is $2.6\times10^{-6}$. Both of these p-values are less than 0.05, the significance level, and so we reject the null hypothesis and say there is evidence against the data following a multivariate normal distribution. 

https://metricgate.com/docs/mardias-test/

These p-values are very small but for extra confidence, especially given our small sample size, we can also do a Henze-Zirkler test with null hypothesis the data follows a multivariate normal distribution. We want independent observations, numeric variables, and it is designed for multiple variables jointly. The Henze-Zirkler test can be for smaller data sets but 30 observations is still small.

```{r}
library(mvnTest)
HZ.test(hydroclean)
```

We get a p-value of 0 from our Henze-Zirkler test and so reject the null hypothesis. So both Mardia's and the Henze-Zirkler test have evidence against multivariate normality. It would be safe to say that our data does not follow a multivariate normal distribution.

https://metricgate.com/docs/henze-zirkler-test/


??
In an attempt to get our data to jointly follow a multivariate normal distribution we could follow in the path of Tanaskovic and team in applying a Box-Cox transformation. If we can get each of our variables to follow a more normal distribution it won't guarantee a joint multivariate normal distribution, but it's worth a look anyhow. The Box-Cox transformation requires our data to be strictly positive, continuous, single-peaked (unimodal), and exhibiting a variance that changes alongside the size of measurements (heteroscedasticity).

```{r}
library(MASS)

trhydroclean <- hydroclean

for (col_name in names(hydroclean)) {
  x <- hydroclean[[col_name]] 
  #Make sure we have only strictly positive values
  if (any(x <= 0, na.rm = TRUE)) {
    x <- x + 0.001 
  }
  
  #Get optimal lambda for the column we are in for its BC transform
  bc <- boxcox(lm(x ~ 1), plotit = FALSE)
  optimal_lambda <- bc$x[which.max(bc$y)]
  
  #Apply the BC transform to this column
  if (abs(optimal_lambda) < 1e-3) {
    trhydroclean[[col_name]] <- log(x)
  } else {
    trhydroclean[[col_name]] <- (x^optimal_lambda - 1) / optimal_lambda
  }
  
}
```

```{r}
#Histograms
par(mfrow = c(2,2))
hist(trhydroclean$T, breaks = 25)
hist(trhydroclean$pH, breaks = 25)
hist(trhydroclean$EC, breaks = 25)
hist(trhydroclean$TS, breaks =25)
par(mfrow = c(2,2))
hist(trhydroclean$Ca2., breaks =25)
hist(trhydroclean$Mg2., breaks =25)
hist(trhydroclean$Na., breaks =25)
hist(trhydroclean$K., breaks =25)
par(mfrow = c(2,2))
hist(trhydroclean$Cl., breaks =25)
hist(trhydroclean$SO2.4, breaks =25)
hist(trhydroclean$HCO.3, breaks =25)
hist(trhydroclean$SiO2, breaks =25)

#QQ-plots
par(mfrow = c(2,2))
qqnorm(trhydroclean$T, main = "T")
qqline(trhydroclean$T)
qqnorm(trhydroclean$pH, main = "pH")
qqline(trhydroclean$pH)
qqnorm(trhydroclean$EC, main = "EC")
qqline(trhydroclean$EC)
qqnorm(trhydroclean$TS, main = "TS")
qqline(trhydroclean$TS)
par(mfrow = c(2,2))
qqnorm(trhydroclean$Ca2., main = "Ca2.")
qqline(trhydroclean$Ca2.)
qqnorm(trhydroclean$Mg2., main = "Mg2.")
qqline(trhydroclean$Mg2.)
qqnorm(trhydroclean$Na., main = "Na.")
qqline(trhydroclean$Na.)
qqnorm(trhydroclean$K., main = "K.")
qqline(trhydroclean$K.)
par(mfrow = c(2,2))
qqnorm(trhydroclean$Cl., main = "Cl.")
qqline(trhydroclean$Cl.)
qqnorm(trhydroclean$SO2.4, main = "SO2.4")
qqline(trhydroclean$SO2.4)
qqnorm(trhydroclean$HCO.3, main = "HCO.3")
qqline(trhydroclean$HCO.3)
qqnorm(trhydroclean$SiO2, main = "SiO2")
qqline(trhydroclean$SiO2)

#The output of the code above can be reduced by par(mfrow = c(3,4))

#Shapiro-Wilk Tests
shapiro.test(trhydroclean$T)
shapiro.test(trhydroclean$pH)
shapiro.test(trhydroclean$EC)
shapiro.test(trhydroclean$TS)
shapiro.test(trhydroclean$Ca2.)
shapiro.test(trhydroclean$Mg2.)
shapiro.test(trhydroclean$Na.)
shapiro.test(trhydroclean$K.)
shapiro.test(trhydroclean$Cl.)
shapiro.test(trhydroclean$SO4_2.)
shapiro.test(trhydroclean$HCO3_.)
shapiro.test(trhydroclean$SiO2)
```
This transformation has definitely improved the look of our histograms and QQ-plots. The right skewness of the histograms have been improved and they look much more symmetric. The QQ-plots also follow the line better now too. Our Shapiro-Wilks test confirm our transformation. All of our p-values have increased to above our 0.05 significance level and so we fail to reject the null hypothesis and conclude that each of our transformed variables follow a normal distribution.

Now we can try our multivariate normal tests on our transformed variables. 
```{r}
library(psych)
library(mvnTest)
mardia(trhydroclean)
HZ.test(trhydroclean)
```
First Mardia's test, we get a p-value of 0.79 for skewness and 0.038 for kurtosis. The skewness p-value is bigger than the significance value 0.05, but because we still reject for the kurtosis component (0.038<0.05) there is still sufficient evidence to conclude our data doesn't follow a multivariate normal distribution.

The Henze-Zirkler test also sees an increased p-value for the transformed data. However the p-value: 0.014, is still smaller than the significance level 0.05 so we still reject the null hypothesis and conclude that the data does not jointly follow a multivariate normal distribution.

The increase in p-values for each test tells us that after the transformation of our variables to be more normally distributed, the joint distribution also becomes closer to a multivariate normal distribution even if it doesn't meet the criteria for Mardia's and Henze-Zirkler tests. 



