---
title: "A Multivariate Analysis on Tanaskovi’c et al. (2012)"
author: "STAT394 Group 6"
editor: visual
format: 
  revealjs:
    incremental: true
    slide-number: true
---

```{r load-libraries}
library(rstatix)
library(MASS)
library(car)
library(here)
library(reshape2)
library(ggplot2)
library(ggcorrplot)
library(GGally)
library(ggbiplot)
library(psych)
library(ggnewscale)
library(pander)
library(tidyverse)
library(dendextend)

set.seed(123)
```

```{r clean-data}
data_spa_waters <- read.csv(file.path("..", "..","Data","Hydrochemical Data.csv"), header=TRUE)
data_spa_waters_classes <- data_spa_waters[-1, -c(1, 2)]
ncolumns_data_spa_waters_classes <- ncol(data_spa_waters_classes)
```

```{r convert-numeric}

data_spa_waters_classes[-c(ncolumns_data_spa_waters_classes - 1, ncolumns_data_spa_waters_classes)] <-
  lapply(
    data_spa_waters_classes[-c(ncolumns_data_spa_waters_classes - 1, ncolumns_data_spa_waters_classes)],
    function(x) round(as.numeric(x), 3)
  )

# Set Structure and Region as factors
data_spa_waters_classes$Geological.structure <- as.factor(data_spa_waters_classes$Geological.structure)
data_spa_waters_classes$Region <- as.factor(data_spa_waters_classes$Region)
```

```{r raw-data}

data_spa_waters_raw <- data_spa_waters_classes[, -c(ncolumns_data_spa_waters_classes - 1, ncolumns_data_spa_waters_classes)]


```

## Presentation outline

- Summary of Tanaskovi'c et al. (2012), Motivation - Hamish
- Our data on Serbian spa waters - Cathrine
- Results of data analysis:
  - Normality tests - Jack
  - Samples consistency - Hamish
  - Correlations, PCA, Factor, cluster analysis - Viktor
  - Hierarchical Cluster Analysis - Cathrine
- Conclusions - Jack

## Correlogram {.smaller}

```{r correlogram, fig.width=6, fig.height=6}

ggcorrplot(cor(data_spa_waters_raw),
           method = "circle",
           type = "lower",
           lab = TRUE)
```

The highest correlation is between electrical conductivity (EC) and total solids (TS) at 0.96, $Na^+$ and $HCO_3^-$ at 0.877 and and $HCO_3^-$ and $K^+$ at 0.827.

## Pairs plot  {.smaller}

```{r pairs-plot1, fig.width=6, fig.height=6}

ggpairs(data_spa_waters_raw)
```

The variables seem to resemble a truncated normal distribution, like the temperature, pH, EC, TS and $SiO_2$, while others have drastically different distributions, therefore we checked them with tests.

## Boxplots by Geological structure

```{r melt-data}

data_spa_waters_classes.melt <- melt(
  data = data_spa_waters_classes[-ncolumns_data_spa_waters_classes],
  id.vars = "Geological.structure",
  variable.name = "Chemical_Compound"
)
```

```{r boxplot-log, fig.width=9}

ggplot(data = data_spa_waters_classes.melt, aes(x = Chemical_Compound, y = value)) +
  geom_boxplot(aes(col = Chemical_Compound), notch = FALSE) +
  scale_y_log10() +
  facet_grid(. ~ Geological.structure) +
  ylab("Measurements") +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        legend.position = "none")
```

## Boxplots by Region

```{r melt-data-region}

data_spa_waters_classes.melt <- melt(
  data = data_spa_waters_classes[-(ncolumns_data_spa_waters_classes-1)],
  id.vars = "Region",
  variable.name = "Chemical_Compound"
)
```

```{r boxplot-log-region, fig.width=9}

ggplot(data = data_spa_waters_classes.melt, aes(x = Chemical_Compound, y = value)) +
  geom_boxplot(aes(col = Chemical_Compound), notch = FALSE) +
  scale_y_log10() +
  facet_grid(. ~ Region) +
  ylab("Measurements") +
  theme(axis.text.x = element_text(size = 7, angle = 90, hjust = 1),
        legend.position = "none")
```

## PCA {.smaller}

```{r pca-reset-raw, fig.height=3, fig.width=4}

data_spa_waters_raw <- data_spa_waters_classes[, -c(ncolumns_data_spa_waters_classes - 1, ncolumns_data_spa_waters_classes)]
```

```{r pca-calculate}

PCA.data_spa_waters_raw <- prcomp(data_spa_waters_raw, center = TRUE, scale = TRUE)
```

```{r pca-scree}
plot(PCA.data_spa_waters_raw, type = "l")
```

4 PCs: 74.727% of the cumulative variance in the data, just under three quarters, with PC1 explaining 35.1%, PC2 17.5%, PC3 12,4% and PC4 9.74% of the total variance.

## PCA - Structure

```{r biplot-geological-structure, fig.height=4, fig.width=6}
ggbiplot(PCA.data_spa_waters_raw, obs.scale = 1, var.scale = 1,
         groups = interaction(data_spa_waters_classes$Geological.structure), ellipse = TRUE)
```

## PCA - Region

```{r biplot-geological-region, fig.height=4, fig.width=6}
ggbiplot(PCA.data_spa_waters_raw, obs.scale = 1, var.scale = 1,
         groups = interaction(data_spa_waters_classes$Region), ellipse = TRUE)

```

## Factor analysis {.smaller}

```{r factanal-diagram}

fa.diagram(factanal(data_spa_waters_raw, factors = 4)$loadings, digits = 3)
```

Four factors explain 64.9% of variation in the data. No single factor explains a large portion of variation.

## Factor analysis - with PCA {.smaller}

```{r principal-varimax}

fa_result <- principal(data_spa_waters_raw, nfactors = 4, rotate = "varimax")

# Check variance accounted for by each rotated factor

fa.diagram(principal(data_spa_waters_raw, nfactors = 4, rotate = "varimax")$loadings, digits = 3)
```

Maintains the 74.73% total variance. These loadings are the closest to the ones described in the paper.

## Cluster analysis - Structure

```{r clustering-kmeans geom, fig.height=4, fig.width=6}

# Build the scores data frame
scores_df <- as.data.frame(PCA.data_spa_waters_raw$x)
scores_df$Group <- data_spa_waters_classes$Geological.structure #Region
scores_df$Label <- 1:nrow(scores_df)

# Variance explained for axis labels
var_explained <- (PCA.data_spa_waters_raw$sdev^2 / sum(PCA.data_spa_waters_raw$sdev^2)) * 100

# Run clustering on the PC scores
k <- 2
km <- kmeans(scores_df[, c("PC1", "PC2")], centers = k, nstart=20)
scores_df$Cluster <- factor(km$cluster)

# Dynamic color/shape mapping for Group
groups <- unique(scores_df$Group)
my_colors <- setNames(c("magenta", "blue", "black", "red", "darkgreen", "orange")[1:length(groups)], groups)
my_shapes <- setNames(c(8, 16, 0, 2, 17, 15)[1:length(groups)], groups)

ggplot(scores_df, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(group = Cluster), type = "norm", level = 0.95,
               color = "grey30", linetype = 2, linewidth = 0.8) +
  geom_point(aes(color = Group, shape = Group), size = 3, stroke = 1.2) +
  geom_text(aes(label = Label, color = Group), vjust = -0.8, hjust = 0.5,
            size = 3.5, fontface = "bold", show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = my_colors) +
  scale_shape_manual(values = my_shapes) +
  labs(x = paste0("Factor 1 (PC1) : ", round(var_explained[1], 2), "%"),
       y = paste0("Factor 2 (PC2) : ", round(var_explained[2], 2), "%")) +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = c(0.2, 0.85),
        legend.background = element_rect(color = "black", linewidth = 0.3))
```

## Cluster analysis - Region

```{r clustering-kmeans region, fig.height=4, fig.width=6}

# Build the scores data frame
scores_df <- as.data.frame(PCA.data_spa_waters_raw$x)
scores_df$Group <- data_spa_waters_classes$Region
scores_df$Label <- 1:nrow(scores_df)

# Variance explained for axis labels
var_explained <- (PCA.data_spa_waters_raw$sdev^2 / sum(PCA.data_spa_waters_raw$sdev^2)) * 100

# Run clustering on the PC scores 
k <- 2  
km <- kmeans(scores_df[, c("PC1", "PC2")], centers = k, nstart=20)
scores_df$Cluster <- factor(km$cluster)

# Dynamic color/shape mapping for Group
groups <- unique(scores_df$Group)
my_colors <- setNames(c("magenta", "blue", "black", "red", "darkgreen", "orange")[1:length(groups)], groups)
my_shapes <- setNames(c(8, 16, 0, 2, 17, 15)[1:length(groups)], groups)

ggplot(scores_df, aes(x = PC1, y = PC2)) +
  stat_ellipse(aes(group = Cluster), type = "norm", level = 0.95,
               color = "grey30", linetype = 2, linewidth = 0.8) +
  geom_point(aes(color = Group, shape = Group), size = 3, stroke = 1.2) +
  geom_text(aes(label = Label, color = Group), vjust = -0.8, hjust = 0.5,
            size = 3.5, fontface = "bold", show.legend = FALSE) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  scale_color_manual(values = my_colors) +
  scale_shape_manual(values = my_shapes) +
  labs(x = paste0("Factor 1 (PC1) : ", round(var_explained[1], 2), "%"),
       y = paste0("Factor 2 (PC2) : ", round(var_explained[2], 2), "%")) +
  theme_bw() +
  theme(legend.title = element_blank(),
        legend.position = c(0.2, 0.85),
        legend.background = element_rect(color = "black", linewidth = 0.3))
```
